---
title: "Choice Experiment Analysis Pipeline"
subtitle: "Exercise choice vs control — school-level behavioural statistics"
author: "B3 Cohort Analysis"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    theme: cosmo
    code-fold: true
    code-summary: "Show code"
    self-contained: true
    fig-width: 8
    fig-height: 5
  pdf:
    toc: true
    number-sections: true
    geometry: margin=2.5cm
    fontsize: 11pt
execute:
  echo: true
  warning: false
  message: false
  eval: false
---

```{r setup}
#| eval: true
#| include: false
# This document is a living overview of the pipeline.
# Set eval: true on individual chunks only when the corresponding
# STEP outputs exist in the expected directories.
knitr::opts_chunk$set(comment = "#>")
```

# Overview

This document describes the complete analysis pipeline for the **exercise-choice experiment** (B3 cohort). Two treatment groups — *exercise choice* and *control* — are compared across a set of school-level behavioural indicators derived from idtracker.ai video-tracking data.

The pipeline is implemented as a series of R scripts run via the master file `00_master_pipeline_choice_exp.R`. Each step produces time-stamped output directories so that any step can be re-run without overwriting prior results.

**Experimental design snapshot:**

| Feature | Value |
|---|---|
| Treatments | `control`, `exercise choice` |
| Housing-tank densities | 4, 8, 12, 16 fish/tank (RE candidate only) |
| Arena fish count | Always 5 |
| Timepoints per trial | 3 × 20-min segments |
| Tracking software | idtracker.ai |
| Analysis unit | Trial × Timepoint (school-level) |

: Experimental design {tbl-cap-location="bottom"}

---

# Pipeline Steps

## Step 1 — Raw trajectory parsing (`idtracker_STEP1_choice_exp.R`)

Reads raw idtracker.ai output (per-session folder with trajectory CSVs), assigns each frame-level detection to one of two polygon hierarchies:

- **Main zone** (`flow` vs `calm`) — equal-area halves of the arena separated by a flow gradient.
- **Sub-zone** (`high`, `medium`, `low`, `calm`) — four strata graded by current intensity.

Zone polygons are defined in two orientation-specific reference files (one per motor-side orientation):

- `zone_reference_choice_exp_FT.csv` — FT (flow-toward) orientation
- `zone_reference_choice_exp_FD.csv` — FD (flow-diagonal) orientation

Polygon vertices are measured in pixel units in ImageJ from a representative video frame of the correct orientation. Zone assignment uses a vectorised ray-casting point-in-polygon test; sub-zone polygons take priority over the broad "flow" polygon where they are spatially nested.

**Output:** `STEP1_output/STEP1_output_<ts>/master_fish_by_frame.csv`

```{r step1-config}
# CONFIG variables used by STEP 1
# Raw idtracker.ai sessions live outside this repository; see config.R.
# The literal formerly hardcoded here pointed at D:/GOT R SCRIPTS/, a
# different project root - a stale path that never matched this tree.
if (!exists("PIPELINE_DATA_DIR")) source(file.path(PROJECT_ROOT, "config.R"))
FRAME_HEIGHT_PX         <- 1080L          # video resolution (informational)
N_PARALLEL_WORKERS      <- NULL           # NULL → detectCores() - 1, max 8
SKIP_PROCESSED_SESSIONS <- FALSE          # TRUE = incremental re-run
```

---

## Step 2 — Activity summary (`activity_analysis_STEP2_choice_exp.R`)

Builds identity-free, school-level summaries at the **trial × timepoint** resolution from the frame-level trajectory data. Because idtracker.ai cannot maintain persistent individual identities in turbulent-flow arenas, all indicators are computed without relying on fish identity labels.

### Trajectory pre-processing

Applied sequentially to each trial's raw (x, y) coordinate streams:

1. **Gap interpolation** — NaN runs bounded by valid coordinates on both sides are filled by linear interpolation.
2. **Jump detection** — frame-to-frame Euclidean displacements are computed; frames whose incoming displacement exceeds *mean + 10 SD* of the session-wide speed distribution are marked as jumps and replaced by interpolation.
3. **Identity-swap detection** — for each consecutive frame pair and each fish pair (i, j), the ratio *(d_self_i + d_self_j) / (d_cross_i + d_cross_j)* is computed. Pairs with ratio > 1.5 (swap ≥ 50 % more parsimonious than current assignment) are flagged; confirmed swap events (verified by Hungarian assignment when `clue` is available) trigger a coordinate-column swap from event onset onward.

### Indicators computed

| Indicator | Column | Method |
|---|---|---|
| Zone occupancy (main) | `prop_time` (long, `zone_level == "main"`) | Per-frame zone proportion, time-weighted mean |
| Zone occupancy (sub) | `prop_time` (long, `zone_level == "sec"`) | Area-corrected, rescaled to sum to 1 |
| Switches per session | `switches_per_session` | NN-matched flow↔calm crossings, normalised to 20 min |
| Zone flux | `zone_flux_per_session` | Mean absolute frame-to-frame Δ(n_flow/n_total), diagnostic only |
| Active fraction | `prop_active` | Proportion of NN-matched blobs exceeding 1 BL s⁻¹ |

**Outputs:**

- `STEP2_output/STEP2_output_<ts>/trial_activity_summary.csv` — wide format, one row per trial × timepoint
- `STEP2_output/STEP2_output_<ts>/trial_occupancy_long.csv` — long format, one row per trial × timepoint × zone

---

## Step 2b — Group dynamics (`group_dynamics_STEP2b_choice_exp.R`)

Computes five collective-behaviour metrics per trial × timepoint using the `swaRm` package, applied frame-by-frame and aggregated to means. Only frames with ≥ 3 valid fish positions are used.

| Metric | Column | Description |
|---|---|---|
| Mean NND | `mean_nnd_cm` | Mean nearest-neighbour distance (cohesion proxy) |
| Polarisation | `mean_polarisation` | Modulus of mean unit heading vector [0, 1] |
| Mean IID | `mean_iid_cm` | Mean inter-individual distance (all pairs) |
| School area | `mean_hull_area_cm2` | School spread |
| School speed | `mean_centroid_spd_cm` | Per-frame displacement of school centroid |

Pixel measurements converted to centimetres via trial-specific calibration (body-length-derived pixel-to-cm ratio).

**Output:** `STEP2b_output/STEP2b_output_<ts>/group_dynamics_summary.csv`

---

## Step 3 — Zone-switch diagnostics (`zone_switch_diagnostics_choice_exp.R`)

Computes Pearson and Spearman correlations between `switches_per_session` (statistical response) and `zone_flux_per_session` (model-free diagnostic), faceted by treatment. Exports a scatter plot and a one-paragraph results text. Supports use of `switches_per_session` as the primary statistical response.

**Output:** `zone_switch_diagnostics/zone_switch_diagnostics_<ts>/`

---

## Step 4 — Graphs (`activity_analysis_GRAPHS_choice_exp.R`)

Produces editable PowerPoint files with two output sets: **timepoint-resolved** (faceted by segment) and **aggregated** (session means).

- Standard indicator plots: x = treatment, colour + shape = treatment (▲ control, ● exercise choice).
- Zone plots: x = zone, colour = zone, shape = treatment; treatments dodged within each zone.

**Output:** `STEP4_graphs/STEP4_graphs_<ts>/`

---

## Step 5 — Statistics (`activity_analysis_STATS_choice_exp.R`)

The primary statistical analysis. Two modules:

- **Module A** — trial × timepoint (primary inference)
- **Module B** — trial-level (timepoints collapsed, companion/robustness check)

### Configuration

```{r step5-config}
TREATMENT_LEVELS  <- c("control", "exercise choice")
DENSITY_LEVELS    <- c(4L, 8L, 12L, 16L)
DENSITY_AS_FACTOR <- TRUE           # treat density as ordered factor in models
TIMEPOINT_LEVELS  <- 1:3

# KR/Satterthwaite comparison (requires lmerTest + pbkrtest)
# Set TRUE to add KR vs SW comparison table to Word report
# SIMPLIFY_SINGULAR_TO_LM <- FALSE  # TRUE: singular LMMs fall back to lm()
```

### Random-effect candidates

Random effects are specified by design, not selected from data to maximise fit.

**Wide-format responses** (one row per trial × timepoint):

| Label | Formula | Rationale |
|---|---|---|
| `tank` | `(1 \| tank)` | Sub-groups from same holding tank share tank-specific variance |
| `trial` | `(1 \| trial_id)` | Three timepoints from one trial share session conditions |
| `trial_tank` | `(1 \| trial_id) + (1 \| tank)` | Both sources |
| `date` | `(1 \| trial_date)` | Trials run on same day may share day-specific conditions |
| `trial_date_r` | `(1 \| trial_id) + (1 \| trial_date)` | Trial + day |
| `density` | `(1 \| fish_density_f)` | Pre-trial housing density (not arena variable) |

**Long-format responses** (one row per trial × timepoint × zone):

| Label | Formula | Rationale |
|---|---|---|
| `trial` | `(1 \| trial_id)` | Paired zone observations within trial are not independent |
| `tank` | `(1 \| tank)` | Tank-level variance |
| `trial_tank` | `(1 \| trial_id) + (1 \| tank)` | Both |
| `nested_tank` | `(1 \| tank / trial_id)` | Trial nested within tank |

AICc is used as a secondary diagnostic only — to drop RE terms that are unidentifiable for a given response (e.g., `(1|tank)` when fewer than 5 tank levels are present), not to choose between biologically distinct fixed-effect structures.

### Analysis workflow (per indicator)

```
1. Shapiro-Wilk test on lm residuals (normality check).
2. Levene's test (variance homogeneity).
3. If SW p < 0.05: try log1p and sqrt transforms;
   retain transform that best satisfies both tests.
   (Skipped for count/proportion responses — handled by GLMM family.)
4. Fit all RE candidates under REML; rank by AICc;
   drop unidentifiable candidates; retain best-supported structure.
5. Type III ANOVA:
   - Gaussian LMM: Kenward-Roger F (primary) + Satterthwaite (comparison)
   - Beta/Poisson/NB GLMM: Wald χ²
6. If any term p < 0.05: Tukey-adjusted EMMs + compact letter display.
7. Marginal + conditional R² (Nakagawa, via performance::r2()).
8. Pseudoreplication check: parallel tank-level mean LM on same response;
   treatment p-values compared for concordance.
9. Benjamini-Hochberg FDR across primary treatment p-values.
```

### Analysis modules

#### Module A — Trial × Timepoint (primary) {#sec-module-a}

```{r module-a-table}
#| eval: true
#| echo: false
analyses <- data.frame(
  ID    = paste0("A", c(1, "1_agg", 2, "2_agg", 3, "3_agg", 4, "4_agg",
                        "5–6", "7–8", "9–10", "11–12", "13–14", "15–16")),
  Label = c(
    "Main-zone occupancy (flow/calm) × Treatment × Zone × Timepoint",
    "Main-zone occupancy × Treatment × Zone (aggregated)",
    "Sub-zone occupancy (h/m/l/calm) × Treatment × Zone × Timepoint",
    "Sub-zone occupancy × Treatment × Zone (aggregated)",
    "Switches per session × Treatment × Timepoint",
    "Switches per session × Treatment (aggregated)",
    "Active fraction × Treatment × Timepoint",
    "Active fraction × Treatment (aggregated)",
    "Mean NND × Treatment (× Timepoint / aggregated)",
    "Polarisation × Treatment (× Timepoint / aggregated)",
    "Mean IID × Treatment (× Timepoint / aggregated)",
    "School area × Treatment (× Timepoint / aggregated)",
    "School speed × Treatment (× Timepoint / aggregated)",
    "Centroid speed × Treatment (aggregated)"
  ),
  Response  = c(
    "prop_time (beta GLMM)", "prop_time (beta GLMM)",
    "prop_time (Gaussian LMM)", "prop_time (Gaussian LMM)",
    "n_main_switches (Poisson/NB GLMM)", "switches_per_session (LMM)",
    "prop_active (beta GLMM)", "prop_active (beta GLMM)",
    "mean_nnd_cm", "mean_polarisation",
    "mean_iid_cm", "mean_hull_area_cm2",
    "mean_centroid_spd_cm", "mean_centroid_spd_cm"
  )
)
knitr::kable(analyses, col.names = c("ID", "Analysis", "Response / model family"),
             caption = "Module A analyses (A1–A16)")
```

**Fixed-effect structures:**

- **Main-zone occupancy (A1):** `treatment * zone * timepoint_f` (beta GLMM, logit link, glmmTMB)
- **Sub-zone occupancy (A2):** `treatment + zone + timepoint_f + treatment:zone + treatment:timepoint_f + treatment:zone:timepoint_f` (Gaussian LMM; lower-order interactions dropped to avoid near-empty cells in the 4-zone × 3-timepoint × 2-treatment design)
- **Switches (A3):** raw count `n_main_switches` with `offset(log(total_time_s))`, Poisson or NB (selected by Pearson dispersion ratio)
- **Active fraction (A4):** `treatment * timepoint_f` (beta GLMM)
- **Group dynamics (A7–A16):** `treatment * timepoint_f` or `treatment` (Gaussian LMM)

For all timepoint-resolved indicators, both categorical (`timepoint_f`) and continuous (`timepoint`) parameterisations are fitted; the better-supported representation (lower AICc) is retained. Aggregated companions use `treatment` (or `treatment * zone` for zone responses) as the only fixed effect.

Sum-to-zero contrasts applied during fitting so that Type-III main effects are marginal.

#### Module B — Trial-level (companion)

Each indicator averaged across timepoints within `trial_id`. Zone responses retain `treatment * zone`; all others use `treatment` alone. `(1|tank)` retained when ≥ 2 tanks contribute > 1 trial; otherwise plain `lm()` fitted. Same inference, post-hoc, and BH correction as Module A.

### Pseudoreplication check

For main-zone occupancy and active fraction: a parallel **tank-level mean LM** is run on the same response. Treatment p-values from the mixed model and the tank-mean LM are compared; concordance is printed to console and saved to `pseudoreplication_check.csv`.

### Multiple-testing correction

Benjamini-Hochberg FDR applied across the **focal treatment p-values** from the primary Module A analyses. Results saved to `bh_correction_summary.csv`.

### Outputs

```
STEP5_stats/STEP5_stats_<ts>/
├── analysis_ready.csv
├── design_balance.csv
├── pseudoreplication_check.csv
├── bh_correction_summary.csv
├── zone_main_timepoint/          # A1
│   ├── model_summary.txt
│   ├── anova_KR.csv
│   ├── anova_SW.csv              # Satterthwaite comparison
│   ├── cld_treatment_zone.csv
│   ├── contrasts_treatment_zone.csv
│   ├── residual_plot.png
│   └── qq_plot.png
├── zone_main_aggregated/         # A1_agg
├── zone_sec_timepoint/           # A2
├── zone_sec_aggregated/          # A2_agg
├── switches_timepoint/           # A3
├── switches_aggregated/          # A3_agg
├── moving_timepoint/             # A4
├── moving_aggregated/            # A4_agg
├── nnd_timepoint/                # A7
├── nnd_aggregated/               # A8
├── polarisation_timepoint/       # A9
├── polarisation_aggregated/      # A10
├── iid_timepoint/                # A11
├── iid_aggregated/               # A12
├── hull_area_timepoint/          # A13
├── hull_area_aggregated/         # A14
├── centroid_speed_timepoint/     # A15
├── centroid_speed_aggregated/    # A16
└── skinny_graphs/
    ├── skinny_graphs_aggregated.pptx
    ├── skinny_graphs_timepoint.pptx
    └── png_preview/
```

The Word report `analysis_report_choice_exp.docx` (written to the pipeline folder) collects all model summaries, ANOVA tables, post-hoc results, and diagnostic information into a single document.

---

# R Packages

```{r packages}
# Core statistics
library(lme4)       # LMM/GLMM fitting
library(lmerTest)   # Kenward-Roger + Satterthwaite df (optional)
library(pbkrtest)   # Kenward-Roger F tests (optional)
library(glmmTMB)    # Beta GLMM, NB GLMM (optional)
library(MASS)       # glm.nb fallback
library(MuMIn)      # AICc model ranking
library(car)        # Type III ANOVA, Levene's test
library(emmeans)    # Estimated marginal means
library(multcomp)   # CLD (compact letter display)
library(multcompView)
library(performance) # R²m / R²c (optional)

# Data wrangling
library(dplyr)
library(tidyr)
library(readr)
library(readxl)
library(stringr)
library(purrr)
library(data.table)

# Parallelism
library(furrr)
library(future)

# Reporting
library(officer)
library(rvg)
library(flextable)
library(ggplot2)
```

---

# Design Notes

## Why identity-free indicators?

The exercise-choice arena generates turbulent water current. Although idtracker.ai produces accurate per-frame positional data, persistent individual identities cannot be reliably maintained across the full 20-min tracking intervals under these conditions. Because the experimental intervention is experienced collectively (a school-level flow gradient), all downstream analyses use identity-independent indicators. The same pipeline is applied to control-treatment trials for methodological consistency.

## Fish density as random effect

Fish density refers to the **housing-tank** stocking density experienced before trials (the arena always contains 5 fish). It is included as a random-effect candidate — not a fixed effect — so models can absorb variance attributable to pre-trial housing rather than treating density as an experimental manipulation. When tank identity and density are confounded 1:1 (detectable via the design balance check), only one of the two intercepts is retained.

## Timepoint parameterisation

For all timepoint-resolved indicators, two parameterisations are fitted:
- **Categorical** (`timepoint_f`): each segment treated as a distinct level.
- **Continuous** (`timepoint`): linear trend across segments.

The better-supported parameterisation (lower AICc) is retained. This is a parameterisation choice, not a hypothesis choice; the underlying fixed-effect terms remain as specified by the design.

## Pseudoreplication

Tank is the repeated-measures unit (sub-groups are drawn from holding tanks). The primary mixed model accounts for tank-level clustering via the random intercept. As a guard, a parallel tank-level mean LM is run on primary responses and treatment p-values are compared for concordance.

---

# Quick-Start

```{r quickstart}
# 1. Set working directory to anywhere inside the repository, then:
source("paths.R")    # discovers PROJECT_ROOT from the .choice-exp-root sentinel

# 2. Point config.R at the raw data, via environment variable:
#    - CHOICE_EXP_DATA_DIR  (path to raw session folders; Step 1 only)
#    - START_FROM_STEP    (1 = full run; 5 = stats only)

# 3. Source master file
source("00_master_pipeline_choice_exp.R")

# 4. Word report written to:
#    D:/GOT R SCRIPTS/choice R pipeline/analysis_report_choice_exp.docx
```

---

# Session Info

```{r session-info}
#| eval: true
sessionInfo()
```
